Propositions as Sessions, Mechanically

### 2015-04-15 paper/formalisation.tex
Finalised the first complete draft.

The locally nameless cofinite quantification (LNCQ) representation for binders
and free variables was adopted for both GV and CP, following
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
variant prevents variable-capture issues when compared with other first-order
approaches. Alternatively, I could have chosen a higher-order abstract syntax
(HOAS) approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, it is less clear how to handle linearity using this
approach and has the disadvantage of being able to represent terms not valid
in the language being mechanised (so called ``exotic'' terms). Further, a
higher-order approach would prevent the use of the Metatheory library which
provides a number of useful tactics for manipulating typing contexts.

Previous efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to
handle a language containing both linear and non-linear contexts. However, to
stay true to CP/GV, I maintain a single context. Interestingly, it became
clear that the mechanised version of CP required explicit weakening and
contraction to facilitate the reduction rules (\S~\ref{sec:cp}).

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
set of variables from being considered as the binder. In contrast, the
traditional ``exists-fresh'' approach, where the binder is only required to be
fresh within the abstraction's body, does not produce strong enough induction
principles in some cases, e.g. weakening~\cite{Aydemir:2008:EFM}. This
decision impacts upon the premises of the CP/GV typing rules (\S~\ref{sec:cp}
and \S~\ref{sec:gv}), however these changes do not alter the typing of
processes nor the semantics of the translation from GV to CP. An equivalence
between ``exists-fresh'' and cofinite quantification definitions for the
simply-typed $\lambda$-calculus ($\stlc$) is given by
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
result could be derived straightforwardly (but tediously) for the CP/GV
system.

@@

\citeauthor{Park:2014:MMW} describe a technique for removing non-linear
contexts from typing judgements. While their work extended to \fpop it is not
clear how to handle an environment containing both non-linear and linear types
as in CP/GV. I wish to maintain as close a relationship as possible to the
pen-and-paper system presented by \citeauthor{Wadler:2014}, so separating out
the non-linear and linear components (as in \fpop) is not an option at this
stage.

\begin{comment}
\section{Issues}

\begin{itemize}
\item What is required of a Metatheory library for linear type systems?
  Permutation reasoning? Weakening/Contraction for non-linear components?
\item Discuss the different approaches that were taken to maintain close
  correspondence with the paper presentation
\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
  out of the proofs e.g. where weakening is applied in the typing derivations
\item Mention the use of Metatheory and its limitations when mechanising
  linear type systems like GV/CP
\end{itemize}
\end{comment}

@@

Figure~\ref{fig:cll} presents the well-typed process relation for CP as an
inductive type in Coq. Each typing rule in CP is represented as a constructor
of the inductive type \coqe$cp_rule$. The inductive type is indexed by a
process (\coqe$proc$) and an environment (\coqe$penv$) indicated by them
appearing after the colon on the first line. \coqe$forall$ quantifies over
hypotheses in a constructor; everything before the comma. Compound hypotheses
are given labels, such as \key{PER}, which is the default name used to
introduce them into the proof context during theorem proving. Some details
such as environment uniqueness have been elided, but notice the use of LNCQ
form within the hypotheses of the rules with binders. These rules exclude a
finite set of names \coqe$L$ from opening the subprocesses where opening a
process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$ and
corresponds to replacing de Bruijn index 0 with \coqe!x!. Bound variables are
represented using de Bruijn indices and free variables are represented using
an abstract type, \coqe$atom$, provided by the Metatheory library. The rules
with binders specify the type of a bound variable in the constructor rather
than a name.

@@

Ordering in environments is ignored so the development required operations for
handling permutations of environments as shown in the \coqe$cp_rule$
definition. The Metatheory library does not currently support permutating
environments, however some basic lemmata were added as a result of this work,
and custom tactics were developed to automatically solve certain permutation
goals via simple proof search. Even so, there are still many cases where
sequences of primitive lemmata, such as those handling append, commutativity
and transitivity, need to be applied manually. A lot of these cases are
similar but there would not appear to be an advantage in developing a tactic
to do the work. Doing so would create a dependency between a hard-to-debug and
obscure Ltac definition and the proofs which utilise it, becoming a burden to
change. Further, such a development philosophy treats Ltac tactics akin to
macros, a shorthand for expressing an oft-repeated sequence of tactics, and it
is doubtful much can be gained from it unless the execution of Ltac tactics
can be made more transparent.

@@

The \key{RED} hypothesis refers to an inductive type defining the principal
cut reductions and commuting conversions on processes. The inductive cases in
\coqe$proc_sub_red$ were complicated to prove due to the process terms being
composed of multiple constructor instances, differing significantly from the
standard operational semantics typically dealt with in $\lambda$-calculus
formalisms. To mitigate the complexity, the proof for each case was written as
a separate lemma (termed ``reduction lemmata''). The last step in the proof of
\coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by the induction
using the reduction lemmata.

@@

\textbf{Weakening and Contraction.} Initially, an implicit notion of weakening
was chosen so as to be in line with \citeauthor{Wadler:2014}'s
definitions. However, explicit weakening is required to differentiate the
principal cut elimination rules for the interactions between servers and
clients; spawning and garbage collection (weakening). The paper presentation
does not specify where weakening could occur due to its implicit nature; it is
tacitly assumed weakening can always be pushed further up the derivation tree
to a base rule e.g. \coqe$cp_fwd$ rule. The same is true of contraction where
it is defined as a substitution of names on process terms, not a
syntax-directed rule.

\begin{samepage}
Contraction is not supported in the formalisation due to the difficulty in
specifying its principal cut elimination rule.  In order to specify the rule
one requires knowledge of the environment typing process \coqe$P$ with the
rule of the form:
\begin{gather*}
\tm{\nu A.(!0(A).P \mid ?z[[A]].Q)}
\quad \becomes \quad \\
\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

where $\Gamma$ is the assumed environment typing \coqe$P$, $\tm{?z[[A]].Q}$
denotes contraction by $z$ of two channels of type $A$ in $Q$ and
$\key{contract}_{\Gamma} \tm{Q}$ contracts $\tm{Q}$ with the bindings in
$\Gamma$. Unfortunately, as explained above, there is no way for the reduction
relation to obtain the environment $\Gamma$.
\end{samepage}

@@

where \coqe$lc_proc P$ specifies that \coqe$P$ is a \textit{local closure}, in
other words contains no unbound de Bruijn indices and $\permb{0}{1}Q$
indicates the permuting of binders 0 and 1 in process $\tm{Q}$. The typing
rule for \coqe$cp_cut$ does not specify any particular split of the linear
environment in the conclusion; splitting is non-deterministic. So, in order to
prove the above statement, it must be true in \key{WT} that \coqe$Q$ takes the
binder of type \coqe$B$ and not \coqe$P$, hence the local closure assumption
on \coqe$open_proc P x$. Similarly, to prove the converse statement
\coqe$forall x, open_proc R x$ must be assumed. For commuting conversions,
permutating binders is essential in formulating the majority of the rules.

@@

Regarding commuting conversions, many require analysing permutations
containing singleton bindings. Consider the following the Coq encoding of the
commuting conversion for output (cf. Figure~\ref{fig:ccoutput}):
\begin{coq}
Lemma reduce_cc_multi_one:
  forall P Q R (x:atom) A B $\Gamma$ (LCQ: lc_proc Q)
         (WT: $\nu$ A.([B] x.(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
    [B] x.($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
\end{coq}

@@

After inverting \key{WT} and specialising the resulting hypotheses, regarding
the subprocesses, with fresh variables the proof state is as follows:

\begin{tabular}{l}
\begin{coq}
1 subgoal
(* ... other elements of the proof context ... *)
CPR : $\open{0}{y}{R}~\tpcp$ y$\sim$A++$\GD$Q
CPQ : Q $\tpcp$ x$\sim$C++$\GD$Q0
CPP : $\open{0}{z}{\open{1}{y}{P}}$ $\tpcp$ z$\sim$B++$\GD$P0
NIN : y $\notin$ {x} $\cup$ dom $\GD$Q0
PER : Permutation $\Gamma$ ($\GD$P++$\GD$Q)
PER0 : Permutation ($\GD$P++y$\sim$A) (x$\sim$B$\otimes$C++$\GD$P0++$\GD$Q0)
\end{coq}\\ \hline
\begin{coq}
x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$
\end{coq}
\end{tabular}
\\~\\

In order to prove the goal, $\Gamma$ must be replaced with the sublists typing
\coqe!P!, \coqe!Q! and \coqe!R!. Therefore, \coqe!PER0! needs to be modified
such that the left-hand side is $\GD$P and the right-hand side has
\coqe!y$\sim$A!  removed. In other words, the binding of \coqe!y! within
\coqe!$\GD$P0! must be extracted (by \coqe!NIN!, \coqe!y! must be in
\coqe!$\GD$P0!). Since many of the commuting conversion rules required this
sort of reasoning, separating the work out into a tactic appears
beneficial. Indeed, Ltac is well-suited for this type of work, searching the
context and manipulating hypotheses of a particular form. The Ltac script in
Figure~\ref{fig:extract} does the job and after executing
\coqe!extract_bnd y A!, \key{PER0} becomes:
\begin{coq}
Permutation ($\GD$P++y$\sim$A) (x$\sim$B$\otimes$C++E1++y$\sim$A++E2++$\GD$Q0)
\end{coq}
where E1 and E2 are sublists of $\GD$P0.

However, the task of removing the binding from the permutation is not
automated to keep information loss within the context transparent to an
individual proof. The tactic is more modular and general in this form,
permitting composition and permutation of tactic sequences.

@@

The theorem for top-level cut elimination appears in Figure~\ref{fig:cutelim}
where $\becomes^\star$ denotes the reflexive and transitive closure of
$\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.
Unfortunately, the theorem has eluded attempts at mechanised proof so far. The
hypothesis of a well-typed process \coqe$P$ is not sufficient to prove the
theorem. For instance, the \coqe$cp_cut$ constructor requires a
``relatedness'' condition on the subprocesses to ensure that they eventually
operate dually on the bound variable. The condition is required after
induction on \key{WT} yields a goal similar to the following:
\begin{coq}
exists R, ($\nu$A.([B]0.($P_1$ $\mid$ $P_2$) $\mid$ Q)) $\becomes^\star$ R
          /\ (is_cut R -> False)
\end{coq}

@@

Following informal proofs, it seems likely that the proof requires a
termination measure involving the length of the derivation. In hindsight,
however, I do not believe the current encoding is the best for this sort of
reasoning. Formalised efforts for proving cut elimination for other logics
have adopted an approach which permits reasoning on the structure of
derivations represented explicitly as
trees~\cite{Tews:2013,Dawson:2010,Dawson:2002}. Whereas, in the approach taken
here, that information is limited and ephemeral in that it is only available
during the proving of a derivation e.g. a reduction lemma. It is future work
to investigate these and other strategies which will assist in formalising
cut-elimination for CP.

@@

Following the original work by \citeauthor{Wadler:2014}, the translation from
GV to CP requires defining translations on types and terms of GV.

@@

The translation of from GV types to CP propositons is shown in
Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
taking a GV session type to its notional dual in CP, reflecting the semantic
difference between channels in GV and those in CP. For GV, the type refers to
the action on the channel endpoint. For CP, the proposition refers to the
\textit{interface} exposed to a user of the channel. For example, output in GV
corresponds to input in CP; the action on the channel endpoint is to output a
value, and to do this the channel must input the value.

@@

The term translation was initially designed using a monadic approach to
streamline handling of the possible error cases. The coq-ext-lib
library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
definitions for monad, functor and applicative type classes and associated
instances for basic Coq types. It is similar in style to Haskell's support for
these objects. Unfortunately, this definition did not yield a suitable
function to prove the translation well-typed because the function needs to be
able to operate structurally on not only terms, but also the environment and
the processes being constructed.

@@

In order to proceed with proving the goal, it is necessary to show that
$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
application of an appropriate lemma and there are two possible ways to prove
this using plain Coq. One way is to apply the lemma to one of \key{PER} or
\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
namely backward reasoning, does not apply since the goal has nothing to do
with the uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert!
tactic one can utilise backward reasoning. An assertion temporarily focuses a
different goal (the asserted goal) which may then be tackled using tactics in
the standard way. Unfortunately, this approach requires the explicit writing
of the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script
cluttering it with unnecessary details and making it sensitive to change
because \coqe!$\GD$P0! and \coqe!$\GD$Q0! are ephemeral names introduced by
theorem prover.

>>>>

The locally nameless cofinite quantification (LNCQ) representation for binders
and free variables was adopted for both GV and CP, following
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
variant prevents variable-capture issues when compared with other first-order
approaches. Alternatively, I could have chosen a higher-order abstract syntax
(HOAS) approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, it is less clear how to handle linearity using this
approach and has the disadvantage of being able to represent terms not valid
in the language being mechanised (so called ``exotic'' terms). Furthermore, a
higher-order approach would prevent the use of the Metatheory library which
provides a number of useful tactics for manipulating typing contexts.

Previous efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to
handle a language containing both linear and non-linear contexts. However, to
stay true to CP/GV, I maintain a single context. \citeauthor{Park:2014:MMW}
describe a technique for removing non-linear contexts from typing
judgements. While their work extended to \fpop it is not clear how to modify
their technique to apply to an environment containing both non-linear and
linear types.

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
set of variables from being considered as the binder. In contrast, the
traditional ``exists-fresh'' approach, where the binder is only required to be
fresh within the abstraction's body, does not produce strong enough induction
principles in some cases, e.g. environment
weakening~\cite{Aydemir:2008:EFM}. This decision impacts upon the premises of
the CP/GV typing rules (\S~\ref{sec:cp} and \S~\ref{sec:gv}), however these
changes do not alter the typing of terms or processes nor the semantics of the
translation from GV to CP. An equivalence between ``exists-fresh'' and
cofinite quantification definitions for the simply-typed $\lambda$-calculus
($\stlc$) is given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM},
and I suspect a similar result could be derived straightforwardly (but
tediously) for the CP/GV system.

@@

@@

Figure~\ref{fig:cll} presents the well-typed process relation for CP as an
inductive type in Coq. Each typing rule in CP is represented as a constructor
of the inductive type \coqe!cp_rule!. The inductive type is indexed by a
process (\coqe!proc!) and an environment (\coqe!penv!) indicated by them
appearing after the colon on the first line. \coqe!forall! quantifies over
hypotheses in a constructor; everything before the comma. Compound hypotheses
are given labels, such as \key{PER}, which is the default name used to
introduce them into the proof context during theorem proving. Some details
such as environment uniqueness have been elided, but LNCQ form within
hypotheses for the rules with binders has been kept. These rules exclude a
finite set of names \coqe!L! from opening the subprocesses where opening a
process \coqe!P! with name \coqe!x! is denoted by \coqe!open_proc P x! and
corresponds to replacing de Bruijn index 0 with \coqe!x!. Bound variables are
represented using de Bruijn indices and free variables are represented using
an abstract type, \coqe!atom!, provided by the Metatheory library. The rules
with binders specify the type of a bound variable in the constructor rather
than a name.

@@

Ordering in environments is ignored so the development required operations for
handling permutations of environments as shown in the \coqe$cp_rule$
definition. The Metatheory library does not currently support permutating
environments, however some basic lemmata were added as a result of this work,
and custom tactics were developed to automatically solve certain permutation
goals via simple proof search. Even so, there are still many cases where
sequences of primitive lemmata, such as those handling append, commutativity
and transitivity, need to be applied manually. A lot of these cases are
similar but it would not be advantageous to develop a tactic to do the
work. Doing so would create a dependency between a hard-to-debug and obscure
Ltac definition and the proofs which utilise it, becoming a burden to
change. Further, such a development philosophy treats Ltac tactics akin to
macros, a shorthand for expressing an oft-repeated sequence of tactics, and it
is doubtful much can be gained from it unless the execution of Ltac tactics
can be made more transparent.

@@

The \key{RED} hypothesis refers to an inductive type defining the principal
cut reductions and commuting conversions on processes. The inductive cases in
\coqe$proc_sub_red$ were complicated to prove due to the process terms being
composed of multiple constructor instances, differing significantly from the
standard operational semantics typically dealt with in $\lambda$-calculus
formalisms. To mitigate the complexity, the proof for each case was written as
a separate lemma (collectively termed ``reduction lemmata''). The last step in
the proof of \coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by
the induction using the reduction lemmata.

@@

\textbf{Weakening and Contraction.} Initially, an implicit notion of weakening
was chosen so as to be in line with \citeauthor{Wadler:2014}'s
definitions. However, explicit weakening is required to differentiate the
principal cut elimination rules for the interactions between servers and
clients; spawning and garbage collection of a server with no clients
(weakening). The paper presentation does not specify where weakening could
occur due to its implicit nature and it is tacitly assumed weakening can
always be pushed further up the derivation tree to a base rule
e.g. \coqe$cp_fwd$ rule. The same is true of contraction where it is defined
as a substitution of names on process terms, not a syntax-directed rule.

\begin{samepage}
Contraction is not supported in the formalisation due to the difficulty in
specifying its principal cut elimination rule.  In order to specify the rule
one requires knowledge of the environment typing process \coqe$P$ with the
rule of the form:
\begin{gather*}
\colored
\tm{\nu \tp{A}.(!0(\tp{A}).P \mid ?z[[\tp{A}]].Q)}
\quad \becomes \quad \\
\colored
\tm{\nu \tp{A}.(!0(\tp{A}).P \mid \nu
                \tp{A}.(!0(\tp{A}).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

where $\Gamma$ is the assumed environment typing $P$,
$\colored\tm{?z[[\tp{A}]].Q}$ denotes contraction by $z$ of two channels of
type $A$ in $Q$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts $\tm{Q}$ with
the bindings in $\Gamma$. Unfortunately, as explained above, there is no way
for the reduction relation to obtain the environment $\Gamma$.
\end{samepage}

@@

where \coqe$lc_proc P$ specifies that \coqe$P$ is a \textit{local closure}, in
other words contains no unbound de Bruijn indices and $\permb{0}{1}Q$
indicates the permuting of binders 0 and 1 in process $\tm{Q}$. The typing
rule for \coqe$cp_cut$ does not specify any particular split of the linear
environment in the conclusion; splitting is non-deterministic. So, in order to
prove the above statement, it must be true in \key{WT} that \coqe$Q$ takes the
binder of type \coqe$B$ and not \coqe$P$, hence the local closure assumption
on \coqe$open_proc P x$. Similarly, to prove the converse statement
\coqe$forall x, open_proc R x$ must be assumed. A different associativity rule
would occur if \coqe!Q! was locally closed in the above. For commuting
conversions, permutating binders is essential in formulating the majority of
the rules.

@@

Regarding commuting conversions, many require analysing permutations
containing singleton bindings. Consider the following Coq encoding of the
commuting conversion for output (cf. Figure~\ref{fig:ccoutput}):
\begin{coq}
Lemma reduce_cc_multi_one:
  forall P Q R (x:atom) A B $\Gamma$ (LCQ: lc_proc Q)
         (WT: $\nu$ A.(x[B].(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
    x[B].($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
\end{coq}

@@

After inverting \key{WT} and specialising the resulting hypotheses regarding
the subprocesses with fresh variables the proof state is as follows:

\begin{tabular}{l}
\begin{coq}
1 subgoal
(* ... other elements of the proof context ... *)
CPR : $\open{0}{y}{R}~\tpcp$ y$\sim$A++$\GD$Q
CPQ : Q $\tpcp$ x$\sim$C++$\GD$Q0
CPP : $\open{0}{z}{\open{1}{y}{P}}$ $\tpcp$ z$\sim$B++$\GD$P0
NIN : y $\notin$ {x} $\cup$ dom $\GD$Q0
PER : Permutation $\Gamma$ ($\GD$P++$\GD$Q)
PER0 : Permutation ($\GD$P++y$\sim$A) (x$\sim$B$\otimes$C++$\GD$P0++$\GD$Q0)
\end{coq}\\ \hline
\begin{coq}
x[B].($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$
\end{coq}
\end{tabular}
\\~\\

In order to prove the goal, $\Gamma$ must be replaced with the sublists typing
\coqe!P!, \coqe!Q! and \coqe!R!. Therefore, \coqe!PER0! needs to be modified
such that the left-hand side is $\GD$P and the right-hand side has
\coqe!y$\sim$A!  removed. In other words, the binding of \coqe!y! within
\coqe!$\GD$P0! must be extracted (by \coqe!NIN!, \coqe!y! must be in
\coqe!$\GD$P0!). Since many of the commuting conversion rules required this
sort of reasoning, separating the work out into a tactic proved
beneficial. Indeed, Ltac is well-suited for this type of work, searching the
context and manipulating hypotheses of a particular form. The Ltac script in
Figure~\ref{fig:extract} does the job and after executing
\coqe!extract_bnd y A!, \key{PER0} becomes:
\begin{coq}
Permutation ($\GD$P++y$\sim$A) (x$\sim$B$\otimes$C++E1++y$\sim$A++E2++$\GD$Q0)
\end{coq}
where E1 and E2 are sublists of $\GD$P0.

However, the task of removing the binding from the permutation is not
automated to keep information loss within the context transparent to an
individual proof. The tactic is more modular and general in this form,
permitting composition and permutation with other tactics.

@@

The theorem for top-level cut elimination appears in Figure~\ref{fig:cutelim}
where $\becomes^\star$ denotes the reflexive and transitive closure of
$\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.
Unfortunately, the theorem has eluded attempts at mechanised proof so far. The
hypothesis of a well-typed process \coqe$P$ is not sufficient to prove the
theorem. For instance, the \coqe$cp_cut$ constructor requires a
``relatedness'' condition on the subprocesses to ensure that they eventually
operate dually on the bound variable. The condition is required after
induction on \key{WT} yields a goal similar to the following:
\begin{coq}
exists R, ($\nu$A.(0[B].($P_1$ $\mid$ $P_2$) $\mid$ Q)) $\becomes^\star$ R
          /\ (is_cut R -> False)
\end{coq}

@@

Following informal proofs, it seems likely that the proof requires a
termination measure involving the length of the derivation. In hindsight
however, I do not believe the current encoding is the best for this sort of
reasoning. Formalised efforts for proving cut elimination for other logics
have adopted an approach which permits reasoning on the structure of
derivations represented explicitly as
trees~\cite{Tews:2013,Dawson:2010,Dawson:2002}. Whereas, in the approach taken
here, that information is limited and ephemeral in that it is only available
during the proving of a derivation e.g. a reduction lemma. It is future work
to investigate these and other strategies which will assist in formalising
cut-elimination for CP.

@@

Following the original work by \citeauthor{Wadler:2014}, the translation from
GV to CP requires defining translations on types, environments and terms of
GV. The translation on environments simply maps the translation on types over
the bindings. The other two translations require more detailed explanation.

@@

The translation of from GV types to CP propositions is shown in
Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
taking a GV session type to its notional dual in CP, reflecting the semantic
difference between channels in GV and those in CP. For GV, the type refers to
the action on the channel endpoint. For CP, the proposition refers to the
\textit{interface} exposed to a user of the channel. For example, output in GV
corresponds to input in CP; the action on the channel endpoint is to output a
value, and to do this the channel must input the value provided by the
programmer.

@@

The term translation was initially designed using a monadic approach to
streamline handling of the error cases. The coq-ext-lib
library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
definitions for monad, functor and applicative type classes and associated
instances for basic Coq types. It is similar in style to Haskell's support for
these objects. Unfortunately, this definition did not yield a suitable
function to prove the translation well-typed because the function needs to be
able to operate structurally on not only terms, but also the environment and
the processes being constructed.

@@

In order to proceed with proving the goal, it is necessary to show that
$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
application of an appropriate lemma and there are two possible ways to prove
this using plain Coq. One way is to apply the lemma to one of \key{PER} or
\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
namely backward reasoning, does not apply since the goal has nothing to do
with the uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert!
tactic one can utilise backward reasoning. An assertion temporarily focuses a
different goal (the asserted goal) which may then be tackled using tactics in
the standard way. Unfortunately, this approach requires the explicit writing
of the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script
cluttering it with unnecessary details and making it sensitive to change
because \coqe!$\GD$P0! and \coqe!$\GD$Q0! are mercurial names introduced by
theorem prover.

### 2015-04-15 paper/futurework.tex
Finalised the first complete draft.

\textbf{Beyond Linear.} Most session-based type systems treat channels as
linear resources which may not be aliased. The ``adoption and focus''
system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
resources.\todo{say what adopt and focus do} The type system requires special
language features, e.g. to manage capabilities. \citeauthor{Fahndrich:2002} do
not provide operational semantics or soundness results for the type system. It
would be interesting to investigate these properties formally in a session
type setting. The ``focus'' rule enforces an invariant on the focussed object
(channel) such that the type is maintained during the focus operation. The
object may be aliased while it is under a focus operation. Such a rule would
appear to only benefit recursive session types. Indeed, focussing on a
non-recursive channel would appear pointless because one is prevented from
communicating across it for the entire scope of the focus. Moreover, it would
be undesirable to permit modification of a session type without guaranteeing
exclusive access to the channel since this could lead to invalid
communication.

>>>>

\textbf{Beyond Linear.} Most session-based type systems treat channels as
linear resources which may not be aliased. The ``adoption and focus''
system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
resources. The ``adopt'' permits aliases on linear resources and ``focus'' is
used to access a linear resource through one of the aliases. The type system
requires special language features to manage
capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
or soundness results for the type system. It would be interesting to
investigate these properties formally in a session type setting. The ``focus''
rule enforces an invariant on the focussed object (channel) such that the type
is maintained during the focus operation. The object may be aliased while it
is under a focus operation. Such a rule would appear to only benefit recursive
session types. Indeed, focussing on a non-recursive channel would appear
pointless because one is prevented from communicating across it for the entire
scope of the focus. Moreover, it would be undesirable to permit modification
of a session type without guaranteeing exclusive access to the channel since
this could lead to invalid communication.
